---
title: "fmri stats"
author: "r poldrack"
date: "February 18, 2015"
output: html_document
---

This notebook walks through the estimation issues involved in processing of fMRI data.

First, we load some necessary libraries:
```{r}

# load utility functions
source('/Users/poldrack/Dropbox/Documents/Teaching/fMRI-stats/fmri_modeling_utilities.R')
library(nlme)
library(knitr)
```

Let's generate some synthetic data, using a blocked design with a single condition that we will compare to the baseline.

```{r}
ntp=360  # number of timepoints, assume 1 timepoint/second for convenience
block_length=30

design=make_boxcar_design(ntp,block_length)

# convolve with hemodynamic response
design_hrf=hrf_conv(design)
# standardize the regressor

design_hrf = (design_hrf - mean(design_hrf))/sd(design_hrf)

plot(design_hrf,type='l')
```

We use discrete cosine basis functions to model low-frequency components of the signal, so let's generate that matrix.

```{r}
dct=get_dct(ntp)
image(t(dct),col = grey(seq(0, 1, length = 256)))
```

Generate the full design matrix and plot it.
```{r}
X=cbind(matrix(t(design_hrf)),dct,matrix(1,nrow=ntp,ncol=1))

image(t(X),col = grey(seq(0, 1, length = 256)))
```

Now we will generate data from the design specified above, using an AR(1)+white noise model. This is not a completely realistic model, but it's good enough to demonstrate the effects of temporal autocorrelation.  First we set the betas.  The first column is the effect size on the design; set this to zero in order to assess null performance.  Columns 2-4 are the low-frequency components, and column 5 is the mean.
```{r}

# betas for parameters in design matrix:
# 1: design regressor
# 2-4: low-frequency components
# 5: intercept
beta=as.matrix(c(0,-10,3,1,1000))
```


Now we loop through generating synthetic data for a number of runs and estimate the parameters using several different models:

```{r}

nruns=100
ar1_param=0.3
estimate_arima=FALSE  # if true, we estimate AR1 param
                      # otherwise, use the true parameter

t=matrix(NA,nrow=nruns,ncol=3)
p=matrix(NA,nrow=nruns,ncol=3)
beta_est=matrix(NA,nrow=nruns,ncol=3)

for (run in 1:nruns) {
	data = generate_ar1wn_data(X,beta,ar1_param=ar1_param)
	
	# 1. fit simple OLS model without trends
	lm.result=lm(data ~ X[,1])
	t[run,1]=summary(lm.result)$coefficients[2,3]
	p[run,1]=summary(lm.result)$coefficients[2,4]
	beta_est[run,1]=lm.result$coefficients[2]
	
	# 2. fit OLS with trends

	lm_trend.result=lm(data ~ X[,1:4])
	t[run,2]=summary(lm_trend.result)$coefficients[2,3]
	p[run,2]=summary(lm_trend.result)$coefficients[2,4]
	beta_est[run,2]=lm_trend.result$coefficients[2]
	
	# 3.  fit GLS using estimated AR structure
	if (estimate_arima) {
    a=arima(lm_trend.result$residual,order=c(1,0,0))
    gls.result=gls(data ~ X[,1:4],correlation=corAR1(a$coef[1]))
  } else {
    gls.result=gls(data ~ X[,1:4],correlation=corAR1(ar1_param))
  }
	t[run,3]=summary(gls.result)$tTable[2,3]
	p[run,3]=summary(gls.result)$tTable[2,4]
	beta_est[run,3]=gls.result$coefficients[2]
		
	}
```

Now print out and plot some of the results.
```{r,echo=FALSE}
summarydata=as.data.frame(rbind(apply(t,2,mean),apply(beta_est,2,mean),apply(beta_est,2,sd)))
rownames(summarydata)=c('mean t','mean beta','sd beta')
names(summarydata)=c('OLS','OLS+detrend','GLS')

kable(summarydata)

par(mfrow=c(3,3))
plot(data)
lines(lm.result$fitted,col='blue')
plot(lm.result$residual,type='l')
acf(lm.result$residual,main='ACF for basic model')

plot(data)
lines(lm_trend.result$fitted,col='blue')
plot(lm_trend.result$residual,type='l')
acf(lm_trend.result$residual,main='ACF for trend model')

plot(data)
lines(gls.result$fitted,col='blue')
plot(gls.result$residual)
acf(gls.result$residual,main='ACF for GLS model')

par(mfrow=c(2,1))
hist(beta_est[,1],20,main=sprintf('beta estimates - OLS (true value = %f)',beta[1]),xlim=c(-1,1),xlab='beta estimate')
lines(c(beta[1],beta[1]),c(0,100),col='blue',lwd=4)
hist(beta_est[,2],20,main=sprintf('beta estimates - OLS + HPF (true value = %f)',beta[1]),xlim=c(-1,1),xlab='beta estimate')
lines(c(beta[1],beta[1]),c(0,100),col='blue',lwd=4)
hist(beta_est[,3],20,main=sprintf('beta estimates - GLS (true value = %f)',beta[1]),xlim=c(-1,1),xlab='beta estimate')
lines(c(beta[1],beta[1]),c(0,100),col='blue',lwd=4)

par(mfrow=c(3,1))
hist(t[,1],20,xlim=c(-5,5),main=sprintf('OLS: mean t=%0.2f',mean(t[,1])))
hist(t[,2],20,xlim=c(-5,5),main=sprintf('OLS + HPF: mean t=%0.2f',mean(t[,2])))
hist(t[,3],20,xlim=c(-5,5),main=sprintf('GLS: mean t=%0.2f',mean(t[,3])))


```

Assess type I error:

```{r}

sig_ct=apply(t>qt(0.95,ntp-1),2,sum)/nruns
sig_ct=as.data.frame(sig_ct)
rownames(sig_ct)=c('OLS','OLS+detrend','GLS')
names(sig_ct)=c('False positive rate')

kable(sig_ct)

```
